---
title: "Air Passenger Demand Forecast"
author: "Mengxuan Lu"
date: "12/11/2020"
output:
  ioslides_presentation: default
  beamer_presentation: default
  slidy_presentation: default
---

```{r global_options}
knitr::opts_chunk$set(fig.path='Figs/')
```

```{r message=FALSE}
library(readr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggpubr)
library(forecast)
library(seasonal)
```
## General Overview
**Goal** \

Forecast the U.S. air passenger demand. 

**Source of the data** \

I downloaded the data from <https://www.transtats.bts.gov/DL_SelectFields.asp?Table_ID=259&DB_Short_Name=Air%20Carriers>.



## Overview the data

```{r message=FALSE, warning=FALSE, paged.print=TRUE}
US_carrier <- read_csv("US_carrier.csv")
glimpse(US_carrier)

```


## EDA
### **Questions I am interested**

1) What are the top 5 busiests airport for departure in 2019?
```{r message= FALSE, warning=FALSE}

top_airport_depart <- US_carrier %>% 
  filter(YEAR == 2019) %>% 
  select(ORIGIN, ORIGIN_CITY_NAME, PASSENGERS) %>% 
  group_by(ORIGIN, ORIGIN_CITY_NAME) %>% 
  summarise(Passengers = sum(PASSENGERS)) %>% 
  arrange(desc(Passengers))

top_airport_depart <- head(top_airport_depart, n = 5)
top_airport_depart <- top_airport_depart %>% 
  separate(ORIGIN_CITY_NAME, c("City", "State"), sep = ",") %>% 
  mutate(Origin_Airport = paste0(ORIGIN,"(", City, ")" )) %>% 
  select(Origin_Airport, Passengers)

p <- ggplot(data = top_airport_depart, 
            aes(x = reorder(Origin_Airport,Passengers), Passengers)) + 
  geom_bar(stat = "identity", width = 0.5, fill = "#FF6666") + 
  scale_y_continuous(limits = c(0,50000000)) +
  coord_flip() 

p + labs(title = "Top 5 busiest airports for departure for 2019", 
         y = "Passengers", x = "Airports") + theme(plot.title = element_text(
           size = 15,face = "bold", hjust = 0.5, vjust = 0.3))
```
2) What are the top 5 busiest airports for arrival in 2019?

```{r message=FALSE, warning=FALSE}
top_airport_arr <- US_carrier %>% 
  filter(YEAR == "2019") %>% 
  select(DEST, DEST_CITY_NAME, PASSENGERS) %>% 
  group_by(DEST, DEST_CITY_NAME) %>% 
  summarise(Passengers = sum(PASSENGERS)) %>% 
  arrange(desc(Passengers))

top_airport_arr <- head(top_airport_arr, n = 5) 

top_airport_arr <- top_airport_arr %>% 
  separate(DEST_CITY_NAME, c("City","State"), sep = ",") %>% 
  mutate(Dest_airport = paste0(DEST, "(", City, ")"))


q <- ggplot(data = top_airport_arr, aes(x = reorder(Dest_airport, Passengers), 
                                   y = Passengers)) +
  geom_bar(stat = "identity", width = 0.5, fill = "#3390FF") +
  scale_y_continuous(limits = c(0, 50000000)) +
  coord_flip()

q + labs(title = "Top 5 busiest airports for arrival for 2019", x = "Airports") +
 theme(plot.title = element_text(size = 15, face = "bold", hjust = 0.5,
                                 vjust = 0.5))
```


```{r message=FALSE, warning=FALSE}
top_airline_year <- US_carrier %>% 
  select(PASSENGERS, UNIQUE_CARRIER, CARRIER_NAME, YEAR) %>% 
  group_by(YEAR, UNIQUE_CARRIER, CARRIER_NAME) %>% 
  summarise(Passengers = sum(PASSENGERS)) %>% 
  arrange(desc(Passengers), UNIQUE_CARRIER) %>% 
  filter(UNIQUE_CARRIER %in% c("WN","DL","AA","UA","OO","US","AS","B6"))

m <- ggplot(data = top_airline_year) +
  scale_x_continuous(breaks = seq(2010, 2020, by = 1)) +
  geom_smooth(aes(x = YEAR, y = Passengers, color = CARRIER_NAME), se = FALSE)

m + labs(title = "The number of passagers from top airlines over year ", 
         x = "Year") +
  theme(plot.title=(element_text(size = 15, face = "bold", hjust = 0.5)),
        legend.title = element_blank()) 
  
```
```{r warning = FALSE, message = FALSE}

top_seat_uti <- US_carrier %>% 
  mutate(EMPTY_SEATS = SEATS - PASSENGERS) %>% 
  filter(SEATS > 0) %>% 
  select(UNIQUE_CARRIER, CARRIER_NAME, EMPTY_SEATS, PASSENGERS, SEATS, YEAR) %>% 
  group_by(YEAR, UNIQUE_CARRIER, CARRIER_NAME) %>% 
  summarise(Total_empty_seats = sum(EMPTY_SEATS), Total_seats = sum(SEATS),
            Passengers = sum(PASSENGERS)) %>% 
  mutate(Seat_utilization = 
                  round(Passengers/Total_seats *100, 2)) %>% 
  arrange(desc(YEAR), desc(Seat_utilization)) %>% 
  select(YEAR, UNIQUE_CARRIER, CARRIER_NAME, Seat_utilization)

seats <- top_seat_uti %>% 
  filter(UNIQUE_CARRIER %in% c("WN","DL","AA","UA","OO","US","AS","B6"))

l <- ggplot(data = seats) +
  geom_smooth(aes(x = YEAR, y = Seat_utilization, 
                  color = CARRIER_NAME), se = FALSE) +
  scale_x_continuous(breaks = seq(2010, 2020, by = 1)) 

l + labs(title = "Seat utilization rate for top airlines", x = "Year", 
         y = "Seat Utiliztion Rate (%)") + 
  theme(plot.title = element_text(size = 15, face = "bold", hjust = 0.5),
        legend.title = element_blank())
  

```

##### RPK = the number of paying passengers x total distance traveled

```{r warning = FALSE, message=FALSE}
# calculate RPK per airline over year, the DISTANCE is in mile, needs to convert to kilometers
Total_RPK <- US_carrier %>% mutate(RPK = PASSENGERS * DISTANCE * 1.61) %>% 
  select(YEAR, UNIQUE_CARRIER, CARRIER_NAME, RPK) %>% 
  group_by(YEAR,UNIQUE_CARRIER, CARRIER_NAME) %>% 
  summarise(TOTAL_RPK = sum(RPK)) %>% 
  arrange(desc(TOTAL_RPK))

RPK <- Total_RPK %>% 
  filter(UNIQUE_CARRIER %in% c("WN","DL","AA","UA","OO","US","AS","B6"))

a <- ggplot(RPK) + 
  geom_smooth(aes(x = YEAR, y = TOTAL_RPK, color = CARRIER_NAME),se=F) +
  scale_x_continuous(breaks = seq(2010, 2020, by = 1)) 
 
a + labs(title = "Revenue Passenger Kilometers(RPK)", x = "Year", y = "RPK") +
  theme(plot.title = element_text(size = 15, hjust = 0.5, face = "bold"), 
        legend.title = element_blank())

```


## Forecast

### **Relation with Covid19 cases?**

```{r warning = FALSE, message=FALSE, figures-side, out.width="50%"}
# import covid19 cases from ourworldindata.org
covid19 <- read.csv("covid-data.csv")

# filter just United States
covid19$date <- ymd(covid19$date, tz = Sys.timezone())
US_covid19 <- covid19 %>% filter(location == "United States") %>% 
  select(date, new_cases) %>% 
  mutate(Year = year(date), Month = month(date), Date = day(date))

# plot the new cases by month 
covid_plot <- ggplot(data = US_covid19, aes(x = date, y =new_cases)) +
  geom_line(color = "blue") +
  scale_x_datetime(date_labels = "%b", date_breaks = "1 month") +
  ggtitle("Covid19 new cases by month") + xlab("Month") + ylab("New cases") +
  theme(plot.title = element_text(hjust = 0.5))

# plot the number of passengers in total in 2020
US_carrier$MONTH <- as.integer(US_carrier$MONTH)
pass_plot <- US_carrier %>% filter(YEAR == 2020) %>% 
  select(MONTH, PASSENGERS) %>% 
  arrange(MONTH) %>% 
  group_by(MONTH) %>% 
  summarise(Passengers = sum(PASSENGERS)) %>% 
  ggplot(aes(x = MONTH, y = Passengers)) + 
  geom_smooth(se = F) + scale_x_discrete(limits = seq(1, 12, by = 1)) +
  ggtitle("Number of Air Passengers by month") + xlab("Month") +
  theme(plot.title = element_text(hjust = 0.5))
 
ggarrange(covid_plot, pass_plot, ncol = 2, nrow = 1) 
```


### Decomposition \

```{r message=FALSE, warning=FALSE}
count_passenger <- US_carrier %>% select(PASSENGERS,YEAR,MONTH) %>% 
  arrange(YEAR, MONTH) %>% 
  group_by(YEAR, MONTH) %>% 
  summarise(passengers = sum(PASSENGERS))

pass <- c(count_passenger$passengers)

# convert to time-series dataset in order to analyze it
passenger = ts(pass, start = 2010, frequency = 12)
fit_dec <- decompose(passenger)
autoplot(fit_dec) + ggtitle("Decompose")

```


```{r}
##(fit.ets <- ets(passenger))

## checkresiduals(fit.ets)

```

### ARIMA
#### auto.arima()
```{r message=FALSE, warning=FALSE}
(fit.arima <- auto.arima(passenger))
checkresiduals(fit.arima)
```

#### Manual ARIMA parameter selection

##### Differencing

```{r message=FALSE, warning=FALSE}
# manual ARIMA model parameter selection
diff_pas <- diff(passenger, lag = 12)
diff_pas %>% ggtsdisplay()
```

##### Parameter selection, choose the lowest AICc
```{r message=FALSE, warning=FALSE}
(myarima <- Arima(passenger, order = c(0,1,1),seasonal = c(0,1,1)))
checkresiduals(myarima)
```

### Compare models ARIMA vs. ETS

```{r message=FALSE, warning=FALSE, echo=TRUE}

mytrain <- window(passenger, start = 2010, end = 2017)
mytest <- window(passenger, start = 2018)

(myarima <- Arima(mytrain, order = c(0,1,1),seasonal = c(0,1,1)))
(fit.ets <- hw(mytrain, seasonal = "additive", damped = TRUE))

fct_ets <- forecast(fit.ets, h = 24) %>% accuracy(passenger)
fct_arima <- forecast(myarima, h = 24) %>% accuracy(passenger)


print("HW ETS model")
fct_ets 

print("ARIMA model")
fct_arima
```

```{r}
##Comparison plot for two models

forecast_ets <- forecast(hw(passenger,damped = TRUE,seasonal = "additive"), h = 12) 
forecast_arima <- forecast(Arima(passenger, order = c(0,1,1),seasonal = c(0,1,1)), 
                           h = 12) 

autoplot(forecast_ets)
autoplot(forecast_arima)


```

